A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2015-10-22 imaging_demo/ms.tex
broke up paragraphs to about 80 character lines

Radio astronomy is entering an era in which interferometers of hundreds to thousands of individual antennas are needed to achieve desired survey speeds. Nowhere is this more apparent than at radio frequencies below 1.4 GHz. The study of the history of hydrogen gas throughout the universe's evolution is pushing technology development towards arrays of low-cost antennas with large fields of view and densely packed apertures. Similarly, the search for transient objects and regular monitoring of the time-dependent sky is driving instruments in the same direction. A number of new telescopes are under development around the world based on this new paradigm, including the Murchison Widefield Array (MWA, \citealt{tin13}), the Precision Array for Probing the Epoch of Reionization (PAPER, \citealt{par10}), the Hydrogen Epoch of Reionization Array (HERA\footnote{http://reionization.org}), the LOw Frequency ARray (LOFAR, \citealt{dev09}), the Canadian Hydrogen Intensity Mapping Experiment (CHIME, \citealt{ban14}), the Long Wavelength Array (LWA, \citealt{ell13}), and the low frequency component of the Square Kilometer Array Low Frequency Aperture Array (SKA-Low \citealt{mel13}).

@@

For certain classes of radio arrays there is an alternative to the FX correlator that can lower the computational burden by directly performing a spatial fast Fourier transform (FFT) on the electric fields measured by each antenna in the array at each time step, removing the cross-correlation step. This relieves the computational scaling from the harsh $\Nant^2$ to the more gentle envelope of $\Npix\log\Npix$, where $\Npix$ is the number of pixels in the Fourier transform \citep[e.g.][]{mor11,teg09,teg10}. This architecture is often referred to as a ``direct imaging" correlator because it eliminates the intermediary cross-correlation data products of the FX and earlier lag correlators, but instead directly forms images from the electric field measurements.

@@

Here we present the first software implementation of the MOFF correlator, and announce the public release of the E-field Parallel Imaging Correlator (EPIC) code. We begin with a technical description of the algorithm in \S \ref{sec:math}, then discuss our particular implementation in \S \ref{sec:software}. We then verify the output data quality from our code in \S \ref{sec:verify} by presenting simulated images from both the EPIC correlator and comparing to a simulated FX correlator. We also demonstrate the performance with real-world data from the LWA. In \S \ref{sec:analysis} we analyze the scaling relationships of the algorithm. We identify specific array design classes where the EPIC correlator is computationally more efficient than the FX algorithm. We conclude and discuss future research prospects in \S\ref{sec:conclusions}.

@@

Electric fields from astrophysical sources, $E(\hat{\mathbf{s}})$, in the sky coordinate system denoted by unit vector $\hat{\mathbf{s}}$ propagate towards the observer as:
\begin{align}
  \widetilde{E}(\mathbf{r}) &= \int E(\hat{\mathbf{s}})\,e^{-i2\pi\mathbf{r}\cdot\hat{\mathbf{s}}}\,\dif^2\hat{\mathbf{s}},
\end{align}
where, $\mathbf{r}$ denotes the observer's coordinate system and $\widetilde{E}(\mathbf{r})$ is the propagated electric field. Thus the propagated electric field is a linear superposition of the electric fields emanating from astronomical sources with appropriate complex phases. It can also be described as a Fourier transform of the electric fields in the sky coordinates.

@@

\begin{align}
  \widetilde{V}_p &= \left\langle \widetilde{E}_a\widetilde{E}_b^\star \right\rangle \label{eqn:cc-vis}\\
                  &= \int\delta(\mathbf{u}-\mathbf{u}_p)\left[\int e^{-i2\pi\mathbf{u}.\hat{\mathbf{s}}}\,B(\hat{\mathbf{s}})\,I(\hat{\mathbf{s}})\,\dif^2\hat{\mathbf{s}}\right] \dif^2\mathbf{u} + n_p,
\end{align}
where, $I(\hat{\mathbf{s}})$ is the sky brightness, $B(\hat{\mathbf{s}})$ is the antenna power response, $\mathbf{u}$ denotes the coordinates of the $uv$-measurement plane, $\mathbf{u}_p\equiv \mathbf{r}_a-\mathbf{r}_b$ is the antenna spacing vector corresponding to the antenna pair $p$, and $n_p$ is the receiver noise. This signifies that the visibility ($\widetilde{V}_p$) measured between a pair of antennas ($p$) is obtained by the multiplying the sky brightness $I(\hat{\mathbf{s}})$ by the antenna power response $B(\hat{\mathbf{s}})$ and Fourier transform from the directional coordinates ($\hat{\mathbf{s}}$) to $uv$ coordinates, which are then sampled at the locations of the antenna spacings (or baselines), namely, $\mathbf{u}_p$ using the sampling function $\delta(\mathbf{u}-\mathbf{u}_p)$ and added to the receiver noise $n_p$. This can be equivalently re-written as:
\begin{align}\label{eqn:software-holography}
  \widetilde{V}_p &= \int\delta(\mathbf{u}^\prime-\mathbf{u}_p)\left[\int \tilde{B}(\mathbf{u}^\prime-\mathbf{u}) \right. \nonumber\\
  &\left. \qquad \times \left[\int e^{-i2\pi\mathbf{u}.\hat{\mathbf{s}}}\,I(\hat{\mathbf{s}})\,\dif^2\hat{\mathbf{s}}\right]\dif^2\mathbf{u}\vphantom{\int}\right] \dif^2\mathbf{u}^\prime + n_p,
\end{align}
where, $\tilde{B}(\mathbf{u})$ denotes the $uv$-space antenna power response obtained by a Fourier transform of $B(\hat{\mathbf{s}})$. Effectively, the multiplication in image space by $B(\hat{\mathbf{s}})$ has been replaced by a convolution with $\tilde{B}(\mathbf{u})$ in $uv$-space. This is the software holographic equivalent of traditional FX correlator output. Following the matrix notation of \citet{mor11}, the above equation can be expressed as:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \widetilde{B}(\mathbf{v},\mathbf{u})\,\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})\,I(\hat{\mathbf{s}}) + \mathbf{n}(\mathbf{v}),
\end{align}
where, the sky brightness $I(\hat{\mathbf{s}})$ is Fourier transformed using $\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})$ and the resultant spatial coherence function is weighted and summed using the antenna power response, $\widetilde{B}(\mathbf{v},\mathbf{u})$ in $uv$-space sampled at the baseline location to obtain the measured visibilities:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \left\langle \widetilde{E}(\mathbf{a})\ast\widetilde{E}(\mathbf{a}^\prime)\right\rangle_t. \label{eqn:matrix-cc-vis}
\end{align}
$\mathbf{m}(\mathbf{v})$ denotes visibilities measured by cross-correlating measured antenna electric fields over all possible pairs of $\mathbf{a}$ and $\mathbf{a}^\prime$. It is the same as equation~\ref{eqn:cc-vis} written in matrix notation.

@@

From equations~\ref{eqn:measured-E-field}, \ref{eqn:software-holography}, \ref{eqn:matrix-cc-vis}, \ref{eqn:antenna-power-breakup} and \ref{eqn:sky-intensity}, the optimal image in equation \ref{eqn:dirty-image-FX} can be expressed equivalently as:
\begin{align}
  I^\prime(\hat{\mathbf{s}}) &= \left\langle \left|\,\mathbf{F}^\textrm{T}(\hat{\mathbf{s}},\mathbf{r})\,\widetilde{\mathbf{W}}^\textrm{T}(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{N}}(\mathbf{a},\mathbf{a})\,\widetilde{E}(\mathbf{a})\,\right|^2\right\rangle_t. \label{eqn:dirty-image-MOFF}
\end{align}
The term inside the angular brackets before squaring has a very similar form as that in equation~\ref{eqn:dirty-image-FX}. It signifies that the measured antenna electric fields are weighted by the antenna noise, weighted and gridded by the antenna aperture kernel, Fourier transformed and finally squared to obtain the same image estimated that would have been obtained using equation~\ref{eqn:dirty-image-FX}.

@@

We have implemented the MOFF imaging technique in our ``E-field Parallel Imaging Correlator'' -- a highly parallelized Object Oriented Python package,\footnote{EPIC package can be accessed at https://github.com/nithyanandan/EPIC} now publicly available. Besides implementing the MOFF imaging algorithm it also includes FX imaging using software holography technique and a simulator for generating electric fields from a sky model.

@@

Figure~\ref{fig:FX-flowchart} shows the flowchart for software holographic imaging from a FX correlator. The antenna-based F-engine is identical to that in the MOFF processing. The electric field spectra each antenna are then cross-multiplied in the X-engine with those from all other antennas to obtain the visibilities $V_\textrm{ij}(f)$ ($\mathbf{m}(\mathbf{v})$ in matrix notation). They are calibrated and time-averaged to obtain $\langle V_\textrm{ij}(f)\rangle$ which are then gridded and imaged to obtain the image $I(f)$. The $I(f)$ obtained from both techniques are identical as explained in \S\ref{sec:math}.
\begin{figure}
  \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{A flowchart of FX imaging in EPIC. The FX process flow shares the F-engine with the MOFF process. Following the F-engine, the electric fields pass through the X-engine to obtain visibilities $V_\textrm{ij}(f)$ which are calibrated and time-averaged. Then they are gridded to obtain the gridded visibilities $V_\textrm{g}(f)$ which are then imaged to obtain the image $I(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

>>>>

Radio astronomy is entering an era in which interferometers of hundreds to
thousands of individual antennas are needed to achieve desired survey speeds.
Nowhere is this more apparent than at radio frequencies below 1.4 GHz. The study
of the history of hydrogen gas throughout the universe's evolution is pushing
technology development towards arrays of low-cost antennas with large fields of
view and densely packed apertures. Similarly, the search for transient objects
and regular monitoring of the time-dependent sky is driving instruments in the
same direction. A number of new telescopes are under development around the world
based on this new paradigm, including the Murchison Widefield Array (MWA,
\citealt{tin13}), the Precision Array for Probing the Epoch of Reionization
(PAPER, \citealt{par10}), the Hydrogen Epoch of Reionization Array (HERA
\footnote{http://reionization.org}), the LOw Frequency ARray (LOFAR,
\citealt{dev09}), the Canadian Hydrogen Intensity Mapping Experiment (CHIME,
\citealt{ban14}), the Long Wavelength Array (LWA, \citealt{ell13}), and the low
frequency component of the Square Kilometer Array Low Frequency Aperture Array
(SKA-Low \citealt{mel13}).

@@

For certain classes of radio arrays there is an alternative to the FX correlator
that can lower the computational burden by directly performing a spatial fast
Fourier transform (FFT) on the electric fields measured by each antenna in the
array at each time step, removing the cross-correlation step. This relieves the
computational scaling from the harsh $\Nant^2$ to the more gentle envelope of
$\Npix\log\Npix$, where $\Npix$ is the number of pixels in the Fourier transform
\citep[e.g.][]{mor11,teg09,teg10}. This architecture is often referred to as a
``direct imaging" correlator because it eliminates the intermediary cross-
correlation data products of the FX and earlier lag correlators, but instead
directly forms images from the electric field measurements.

@@

Here we present the first software implementation of the MOFF correlator, and
announce the public release of the E-field Parallel Imaging Correlator (EPIC)
code. We begin with a technical description of the algorithm in \S
\ref{sec:math}, then discuss our particular implementation in \S
\ref{sec:software}. We then verify the output data quality from our code in \S
\ref{sec:verify} by presenting simulated images from both the EPIC correlator
and comparing to a simulated FX correlator. We also demonstrate the performance
with real-world data from the LWA. In \S \ref{sec:analysis} we analyze the
scaling relationships of the algorithm. We identify specific array design classes
where the EPIC correlator is computationally more efficient than the FX
algorithm. We conclude and discuss future research prospects in \S
\ref{sec:conclusions}.

@@

Electric fields from astrophysical sources, $E(\hat{\mathbf{s}})$, in the sky
coordinate system denoted by unit vector $\hat{\mathbf{s}}$ propagate towards
the observer as:
\begin{align}
  \widetilde{E}(\mathbf{r}) &= \int E(\hat{\mathbf{s}})\,e^{-i2\pi\mathbf{r}\cdot\hat{\mathbf{s}}}\,\dif^2\hat{\mathbf{s}},
\end{align}
where, $\mathbf{r}$ denotes the observer's coordinate system and $\widetilde{E}
(\mathbf{r})$ is the propagated electric field. Thus the propagated electric
field is a linear superposition of the electric fields emanating from
astronomical sources with appropriate complex phases. It can also be described
as a Fourier transform of the electric fields in the sky coordinates.

@@

\begin{align}
  \widetilde{V}_p &= \left\langle \widetilde{E}_a\widetilde{E}_b^\star \right\rangle \label{eqn:cc-vis}\\
                  &= \int\delta(\mathbf{u}-\mathbf{u}_p)\left[\int e^{-i2\pi\mathbf{u}.\hat{\mathbf{s}}}\,B(\hat{\mathbf{s}})\,I(\hat{\mathbf{s}})\,\dif^2\hat{\mathbf{s}}\right] \dif^2\mathbf{u} + n_p,
\end{align}
where, $I(\hat{\mathbf{s}})$ is the sky brightness, $B(\hat{\mathbf{s}})$ is the
antenna power response, $\mathbf{u}$ denotes the coordinates of the $uv$-
measurement plane, $\mathbf{u}_p\equiv \mathbf{r}_a-\mathbf{r}_b$ is the antenna
spacing vector corresponding to the antenna pair $p$, and $n_p$ is the receiver
noise. This signifies that the visibility ($\widetilde{V}_p$) measured between a
pair of antennas ($p$) is obtained by the multiplying the sky brightness
$I(\hat{\mathbf{s}})$ by the antenna power response $B(\hat{\mathbf{s}})$ and
Fourier transform from the directional coordinates ($\hat{\mathbf{s}}$) to $uv$
coordinates, which are then sampled at the locations of the antenna spacings (or
baselines), namely, $\mathbf{u}_p$ using the sampling function
$\delta(\mathbf{u}-\mathbf{u}_p)$ and added to the receiver noise $n_p$. This can
be equivalently re-written as:
\begin{align}\label{eqn:software-holography}
  \widetilde{V}_p &= \int\delta(\mathbf{u}^\prime-\mathbf{u}_p)\left[\int \tilde{B}(\mathbf{u}^\prime-\mathbf{u}) \right. \nonumber\\
  &\left. \qquad \times \left[\int e^{-i2\pi\mathbf{u}.\hat{\mathbf{s}}}\,I(\hat{\mathbf{s}})\,\dif^2\hat{\mathbf{s}}\right]\dif^2\mathbf{u}\vphantom{\int}\right] \dif^2\mathbf{u}^\prime + n_p,
\end{align}
where, $\tilde{B}(\mathbf{u})$ denotes the $uv$-space antenna power response
obtained by a Fourier transform of $B(\hat{\mathbf{s}})$. Effectively, the
multiplication in image space by $B(\hat{\mathbf{s}})$ has been replaced by a
convolution with $\tilde{B}(\mathbf{u})$ in $uv$-space. This is the software
holographic equivalent of traditional FX correlator output. Following the matrix
notation of \citet{mor11}, the above equation can be expressed as:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \widetilde{B}(\mathbf{v},\mathbf{u})\,\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})\,I(\hat{\mathbf{s}}) + \mathbf{n}(\mathbf{v}),
\end{align}
where, the sky brightness $I(\hat{\mathbf{s}})$ is Fourier transformed using
$\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})$ and the resultant spatial coherence
function is weighted and summed using the antenna power response,
$\widetilde{B}(\mathbf{v},\mathbf{u})$ in $uv$-space sampled at the baseline
location to obtain the measured visibilities:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \left\langle \widetilde{E}(\mathbf{a})\ast\widetilde{E}(\mathbf{a}^\prime)\right\rangle_t. \label{eqn:matrix-cc-vis}
\end{align}
$\mathbf{m}(\mathbf{v})$ denotes visibilities measured by cross-correlating
measured antenna electric fields over all possible pairs of $\mathbf{a}$ and
$\mathbf{a}^\prime$. It is the same as equation~\ref{eqn:cc-vis} written in
matrix notation.

@@

From equations~\ref{eqn:measured-E-field}, \ref{eqn:software-holography},
\ref{eqn:matrix-cc-vis}, \ref{eqn:antenna-power-breakup} and
\ref{eqn:sky-intensity}, the optimal image in equation \ref{eqn:dirty-image-FX}
can be expressed equivalently as:
\begin{align}
  I^\prime(\hat{\mathbf{s}}) &= \left\langle \left|\,\mathbf{F}^\textrm{T}(\hat{\mathbf{s}},\mathbf{r})\,\widetilde{\mathbf{W}}^\textrm{T}(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{N}}(\mathbf{a},\mathbf{a})\,\widetilde{E}(\mathbf{a})\,\right|^2\right\rangle_t. \label{eqn:dirty-image-MOFF}
\end{align}
The term inside the angular brackets before squaring has a very similar form as
that in equation~\ref{eqn:dirty-image-FX}. It signifies that the measured antenna
electric fields are weighted by the antenna noise, weighted and gridded by the
antenna aperture kernel, Fourier transformed and finally squared to obtain the
same image estimated that would have been obtained using equation~
\ref{eqn:dirty-image-FX}.

@@

We have implemented the MOFF imaging technique in our ``E-field Parallel Imaging
Correlator'' -- a highly parallelized Object Oriented Python package,
\footnote{EPIC package can be accessed at https://github.com/nithyanandan/EPIC}
now publicly available. Besides implementing the MOFF imaging algorithm it also
includes FX imaging using software holography technique and a simulator for
generating electric fields from a sky model.

@@

Figure~\ref{fig:FX-flowchart} shows the flowchart for software holographic
imaging from a FX correlator. The antenna-based F-engine is identical to that in
the MOFF processing. The electric field spectra each antenna are then cross-
multiplied in the X-engine with those from all other antennas to obtain the
visibilities $V_\textrm{ij}(f)$ ($\mathbf{m}(\mathbf{v})$ in matrix notation).
They are calibrated and time-averaged to obtain $\langle V_\textrm{ij}(f)\rangle$
which are then gridded and imaged to obtain the image $I(f)$. The $I(f)$ obtained
from both techniques are identical as explained in \S\ref{sec:math}.
\begin{figure}
  \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{A flowchart of FX imaging in EPIC. The FX process flow shares the
    F-engine with the MOFF process. Following the F-engine, the electric fields
    pass through the X-engine to obtain visibilities $V_\textrm{ij}(f)$ which are
    calibrated and time-averaged. Then they are gridded to obtain the gridded
    visibilities $V_\textrm{g}(f)$ which are then imaged to obtain the image
    $I(f)$.}
  \label{fig:FX-flowchart}
\end{figure}
